# Data Checks and Processing

```{r database, echo = FALSE, eval=FALSE, warning=FALSE}
library(DBI)
library(RSQLite)
library(tidyverse)
library(lme4)
ms_thesis_db <- dbConnect(RSQLite::SQLite(), "ms_thesis.db") 
aspen_plots <- dbGetQuery(ms_thesis_db, "SELECT * FROM aspen_plots;")
trees <- dbGetQuery(ms_thesis_db, "SELECT * FROM trees;")
measurements <- dbGetQuery(ms_thesis_db, "SELECT * FROM measurements;")
```
## Data checks
Before conducting any analyses I must first check for data entry errors and refine 
the data to exclude temporary minimum height placeholders to ensure I am only working
with actual measurements.

### Aspen Plots
```{r aspen_plot_eval,echo = TRUE, eval=FALSE, warning=FALSE}
anyNA(aspen_plots)
which(is.na(aspen_plots)) 
range(aspen_plots$utm_x_start) 
range(aspen_plots$utm_y_start)
range(aspen_plots$utm_x_end, na.rm = TRUE)
range(aspen_plots$utm_y_end, na.rm = TRUE)
length(unique(aspen_plots$plot_id))
```
Check table aspen_plots. `anyNA` was used to check for NA's, `range` was used to
determine if I mixed up coordinates, and `length` and `unique` were used to endure 
all plots were accounted for. I ended up needing to fix some coordinates that had
the latitude and longitude swapped.

### Trees
```{r trees_eval, echo = TRUE, eval=FALSE, warning=FALSE}
anyNA(trees)
range(trees$tree_id)
range(trees$tree_number)
range(trees$plot_id)
range(trees$year)
```
Similar to before, I checked the data for errors. I found that tree numbers for 
2020 were not put into the .csv, rectified the original .csv file.

### Measurements
```{r measurements_eval, echo = TRUE, eval=FALSE, warning=FALSE}
range(measurements$height, na.rm = TRUE)
range(measurements$cag, na.rm = TRUE)
range(measurements$dbh, na.rm = TRUE)
unique(measurements$winter_browse, na.rm = TRUE)
unique(measurements$summer_browse,na.rm = TRUE)
unique(measurements$new_sucker, na.rm = TRUE)
sort(unique(tall_ya$height), decreasing  = TRUE) 
flitered_measurements <- filter(measurements, height < 550) %>% 
  filter(!(height %in% c(300,350,400,450, 500, 550)))
```
In our data, we use minimum estimates for heights that we can't measure. I want to
filter those out as they are not actual measurements, just estimates. These heights are 
300, 350, 400, 450, 500, 550. I will also remove all measurements over 550 as 
those are either estimates, or generated from a dbh formula as it is not possible 
to measure a 550cm tree with a meter stick. To do this, I selected all measurements
under 550 cm, and then selected the opposite including the placeholders above. 
This removed 1554 rows which had minimums.

## Create a table to use for graphs
```{r join, echo = TRUE, eval=FALSE}
young_aspen <- flitered_measurements %>% 
  left_join(trees, by = "tree_id") %>% 
  select(height, cag, year, plot_id, tree_id, winter_browse)
write.csv(young_aspen, file = "young_aspen.csv", row.names = FALSE)
```
Here I join my filtered measurements to the trees table and select for the columns
height, cag, year, plot_id, tree_id, and winter_browse. I saved this table as a .csv
to use when graphically explore my data in the next chapter.
